---
title: "Simulation"
author: "Dr. Ibrahim"
date: "2026-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls())
```

```{r}
library(mice)
```


#  Read the  dataset
```{r}
d.cd4<-read.csv("d.cd4.csv")
ord<-order(d.cd4$patient_id, d.cd4$days.art)
d.cd4<-d.cd4[ord,]
head(d.cd4)
```


```{r}
d.cd4$mode1 <- d.cd4$mode
d.cd4$mode1[d.cd4$mode == "unknown"] <- NA
```

# distribution of each category
```{r}
table(d.cd4$mode1, useNA = "ifany")
```


# percentage distribution of each category
```{r}
prop.table(table(d.cd4$mode1, useNA = "ifany")) * 100
```


# created a reduced dataset containing only the variables needed for the analysis and imputation
```{r}
d.cd4.reduced <- d.cd4[, c("patient_id","country","days.art",
                          "male_y","mode1","age","class",
                          "cd4.baseline","cd4_v")]
```

# converted mode1 into a factor variable
```{r}
d.cd4.reduced$mode1 <- as.factor(d.cd4.reduced$mode1)
```




# complete case analysis
```{r}
mod.gee.b <- geeglm(cd4_v ~ days.art + mode1 + male_y + age + 
                    class*days.art + cd4.baseline,
                    family="gaussian",
                    id=patient_id,
                    data=d.cd4.reduced,
                    subset=(country=="Brazil" & !is.na(mode1)),
                    corstr="exchangeable")
```


```{r}
mod.gee.b
```


# Multiple Imputation
```{r}

Nimps <- 10

d.imputed <- mice(d.cd4.reduced,
                  method = "pmm",
                  m = Nimps,
                  seed = 10)
```


# GEE model on each of the imputed datasets
```{r}
mod.gee.i <- with(d.imputed,
                  geeglm(cd4_v ~ days.art + male_y + mode1 + age +
                         class*days.art + cd4.baseline,
                         family="gaussian",
                         id=patient_id,
                         subset=country=="Brazil",
                         corstr="exchangeable"))
```



# combined the results (Rubin’s Rule) which combines estimates and adjusts standard errors
```{r}
pooled_results <- pool(mod.gee.i)
pooled_results
```

# flat dataset, where each patient appears only once instead of multiple times.
```{r}
d.cd4.flat1<-d.cd4[!duplicated(d.cd4$patient_id),
                   c("patient_id","country","days.art","male_y","mode1","age","class","cd4.baseline","cd4_v")]
d.cd4.flat1$mode1<-as.factor(d.cd4.flat1$mode1)

```





# wide dataset
```{r}
d.imp1 <- mice(d.cd4.flat1, method="pmm", m=Nimps, seed=10)
```



# CD4 measurements over time
```{r}
d.cd4.small <- d.cd4[, c("patient_id","country","days.art","cd4_v")]
```




# matrices to stores the regression coefficients, and variances
```{r}
ests <- vars <- matrix(NA, 
                       nrow = length(mod.gee.b$coefficients), 
                       ncol = Nimps)
```


```{r}
#loop through each of the 10 imputed datasets. 
for (i in 1:Nimps) {
  d1 <- complete(d.imp1, action = i)

  # removed variables that are not needed from the imputed dataset to avoid duplication when merging.This ensures that we only retain the patient-level variables from the imputation  
  
  d1$cd4_v <- d1$country <- d1$days.art <- NULL
  
  
  #Next, I merged the imputed patient-level dataset with the longitudinal CD4 dataset.

#This reconstructs the full dataset with repeated CD4 measurements and imputed covariates.”
    dnext1 <- merge(d.cd4.small, d1, by="patient_id")
    
    #fit the same GEE model on the reconstructed dataset for each imputation.

      mod.gee.imp <- geeglm(cd4_v ~ days.art + male_y + mode1 + age + 
                        class*days.art + cd4.baseline, 
                        family="gaussian",
                        id=patient_id,
                        subset=country=="Brazil",
                        data=dnext1,
                        corstr="exchangeable")
      

# stored the regression coefficients and their variances for each imputed dataset. 
        ests[,i] <- mod.gee.imp$coefficients
  vars[,i] <- summary(mod.gee.imp)$coefficients[,2]^2
}
    
```

# calculated the final pooled estimates by taking the average of the coefficients across all imputations.

```{r}
apply(ests, 1, mean)
sqrt(apply(vars, 1, mean) + (1 + 1/Nimps) * apply(ests, 1, var))

```


#Finally, I compared the results from the wide imputation approach with the earlier long-format imputation results.

```{r}
pooled_results
```

